Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2017-03-28 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Replace fig Ch 9   In corporated revised to Ch 1 and 2

This section reviews the fire problem to demonstrate the need and value to society that fire investigation can provide. Following the review, a brief presentation on the analysis  of fire patterns, the use of analytical tools, and the current challenges to the field of fire forensics is given as they are the reason for this study.  At the end of this chapter, an example of a fire incident is presented. In Chapter 9, we will return to the example again.

\subsection{The Need for Fire Investigation}
Unwanted fires result in loss of life, property, and can create an adverse economic impact on a community.  Recent findings from an archaeological site in South Africa, indicate that mankind has productively used fire for approximately 1 million years~\cite{Berna:2012}.  This is 300 thousand years earlier than previously thought.  Despite mankind's extensive history with fire, according to the World Health Organization, an estimated 265,000 people die each year as a result of burn injuries~\cite{WHO:2014}.  Based on an estimated global population of 7.3 billion people, the death rate due to burns from all sources would be 3.6 per 100,000 people.  The World Fire Statistics Center collects data from European, North American and a few Asia-Oceania countries.  Within this group of countries the fire deaths per 100,000 population range from 0.02 (Singapore) to 2.03 (Finland)~\cite{Climate:2014}. There are a significant number of factors that determine the variability of fire death rates between nations, socio-economic conditions being a leading factor~\cite{WHO:2014}.

In the United States, the loss of life due to unwanted fire has been decreasing steadily since the 1970’s.  This trend does not include loss of life as a result of the terrorist attack on September 11, 2001.  The estimated number of fire deaths in the U.S. in 1975 was 8,100, or 3.75 deaths per 100,000 population~\cite{America_Burning_Revisited}.  In 2013, there were 3,240 civilian fire fatalities, or 1.02 deaths per 100,000 population~\cite{Karter:2014}.  During this same period, the number of fires in the United States has decreased from 2.5 million fires to 1.24 million fires.  There are many reasons that led to this decrease in loss of life due to fire in the United States.  One example is research and data that enabled the development of product safety standards, improved building codes, improved fire safety systems, and fire prevention programs.  None of this happened without the constant effort of agencies at the Federal, state, and, local levels working with professional fire organizations in order to deliver the five E’s of fire prevention; Engineering, Education, Enforcement, Economic Incentive, and Emergency Response~\cite{FEMA:2013}.

@@

According to the NFPA, there were approximately 282,600 intentionally set fires each year during the period of 2007 through 2011 in the United States. The average annual loss totals for that period were approximately 420 civilian deaths and 1,360 civilian injuries. More than \$1.3 billion dollars (U.S.) was lost per year due to direct property damage. Structure fires represent less than 20\% of intentionally set fires.  However, fires involving structures account for more than 84\% of the civilian deaths, injuries and property loss from all intentional fires.  Approximately two-thirds of the intentional structure fires occurred in occupied and operating residential occupancies.  As a result, 95\% of the civilian deaths and 86\% of the civilian injuries which are attributed to intentional fires occurred in residential occupancies~\cite{Campbell:2014}.

A subset of the intentional fires are considered arson.  The U.S. Federal Bureau of Investigation (FBI) Uniform Crime Reporting (UCR) Program defines arson as any willful or malicious burning or attempting to burn, with or without intent to defraud, a dwelling house, public building, motor vehicle or aircraft, personal property of another, etc.~\cite{Crime:2010}.  The national rate of arson fire cases cleared by arrest or exceptional means is less than 20\%.  Of those cases approximately 46\% of the suspects were prosecuted~\cite{Campbell:2014}.  Hence less than 10\% of the arson cases resulted in successful prosecution.

The statistics cited above are believed to be the best available in the United States, which is perhaps one of the better fire incident databases in the world.  However, an investigative report by Icove and Hargrove indicates that arson fires are underreported to the U.S. National Fire Incident Reporting System (NFIRS).  The investigation provided examples from several major cities, such as New York City. In 2011, 11 arson fires in buildings were reported to NFIRS, while the Fire Department of the City of New York's Bureau of Fire Investigation recorded 1,347 arson fires in buildings that year.  Similar discrepancies were found in eight other large U.S. cities.  On average, three-fourths of the arsons uncovered by city fire investigators went unreported to NFIRS~\cite{Icove_2014}.  In other words, the number of arson related incidents is likely to be under represented in the fire incident data.

\subsection{The Use of Fire Patterns in Fire Investigation}
Fires are investigated in order to determine the ``origin and cause'' of the fire.  In cases of arson, there is an additional effort to determine who was responsible for setting the fire.  This investigation process must follow the scientific method, as documented in NFPA 921~\cite{NFPA:921}.  There are numerous textbooks on the subject of conducting a fire investigation~\cite{Almirall:2004,Fire_Investigation,DeHaan:2012,Icove:2013,Lentini:2006,Noon:1995}. The NFPA guide and the texts provide information how to collect data from the scene, analyze the data, and develop a hypothesis about the fire.

Patterns produced by the fire are in many cases a significant portion of the data collected and are analyzed at the scene to determine the area of origin.  One of the basic methods of documenting the fire scene is to photograph fire patterns.  As noted by Icove, DeHaan, and Haynes  ``'the ability to document and interpret fire patterns accurately is essential to investigators reconstructing fire scenes…''~\cite{Icove:2013}.  A fire pattern is defined in NFPA~921 as, ``the visible or measurable physical change, or identifiable shapes, formed by a fire effect or group of fire effects''~\cite{NFPA:921}.

DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
\item Surface deposits -– no irreversible effect on surface
\item Surface thermal effects –- physical change such a discoloration or melting
\item Charring -– evidence of surface burning
\item Penetration -– charring below the surface
\item Consumption –- loss of surface material, charring throughout
\end{enumerate}
Lentini points out that most fire patterns are generated and later observed on two-dimensional surfaces at the place where those surfaces intersect with the three-dimensional fire. Various types of fire patterns, such as; ``V-shaped'', ``hour-glass'', and ``inverted cone'', have come from common observation at actual fire scenes~\cite{Lentini:2006}.  As a result, the observations are typically qualitative in nature.

Previous fire pattern research by the National Institute of Justice (NIJ), the National Institute of Standards and Technology (NIST), and the United States Fire Administration (USFA) has shown that fire patterns provide data useful for the determination of fire origin.  The reports noted the impact of ventilation on the development of the burn patterns~\cite{Shanley:1997,Putorti:1997}. A large number of other factors affect the formation of these patterns: burn time, heat release rate of the fire source, fire exposure, target fuel composition, adjacent fuel(s), and compartmentalization, to name a few. Given the limited number of experiments in the literature and the large number of variables, it has been difficult to develop a cause and effect relationship between the fire scenarios and the resulting patterns, based solely on the research.

\subsection{The Use of Fire Models in Fire Investigation}
Fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigator’s hypothesis on origin and cause~\cite{Sutula:2001}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs)~\cite{NUREG_1805}. These spreadsheets incorporate references to the source documents and experimental work upon which the correlations are based. NRC tasked NIST with conducting a verification and validation study of these correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study, because a sufficient amount of reliable data was not available to conduct a proper validation.  The correlations for flame height and time to ignition were not included in the validation study.

@@

Of the active field models, FDS version 6 has one of the most rigorous and publicly accessible verification and validation programs~\cite{FDS_Verification_Guide,FDS_Validation_Guide}. More than 45 experimental data sets are used to quantify the prediction capabilities of the model.  Results are presented in terms of model bias and variation to inform users of whether particular quantities (temperature, heat flux, pressure, etc.) are over or under predicted compared to experimental data.

In order for fire models to be demonstrated as reliable and suitable for use in court, studies such as the one conducted here need to be conducted to provide the scientific basis for the use of fire models for re-construction.  These research results may enable fire investigation teams to have a better understanding of the fire scene and enable improved support of their hypothesis on the area of origin based on fire patterns.  This idea was discussed when the Fire Protection Research Foundation convened a Research Advisory Council on Post-fire Analysis in 2002.  Recommendations for research and development in their white paper included, ``advance the capabilities of computational fluid dynamics fire modeling, particularly as applied to fire ignition scenarios and fire pattern development and interpretation''~\cite{RAC:2002}.

Since fire pattern interpretation and analysis is a critical step in most fire investigations, it is surprising to find the limited amount of research and analysis tools directly related to this practice.  Gorbett recently conducted a review of fire pattern research that has been conducted over the last 80 years. The review points out many gaps, especially with how fire patterns are used to determine the area of origin.  A key limitation is the assumption by many of the researchers, or text authors, that investigators are able to visibly assess varying degrees of fire damage and then determine the area of origin~\cite{Gorbett_2015}. Studies have shown that investigators have not been able to use only the location of the fire patterns as means of reliably determining the area of origin.  Some amount of analysis and fire dynamics knowledge is required~\cite{Carmen_2008},~\cite{Tinsley_2013}.

\subsection{Fire Investigation Under Review}
The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA~921, {\em Guide for Fire and Explosion Investigations}~\cite{NFPA:921}.  The first edition of NFPA~921 was issued in 1992.  As provided by the document scope, it ``is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' However, NFPA~921 can only incorporate the fire science and research results that have been produced.  Although NFPA~921 is a ``guide'' as opposed to a ``standard,'' it is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.

As the result of the review of several arson cases, most notably the cases of Ernest Ray Willis and Cameron Todd Willingham, the practice of fire investigation as a forensic science was called into question.  Beyler conducted a review of both cases for the Texas Forensic Science Commission~\cite{Beyler:2009}.  Both fires occurred prior to the release of NFPA~921 in 1992.  The Beyler analysis, issued in August of 2009, concluded that the investigations of both cases ``did not comport with either the modern standard of care expressed by NFPA 921, or the standard of care expressed by fire investigation texts and papers in the period 1980 {\textendash} 1992.'' The fire investigators had a poor understanding of science and their findings of arson could not be sustained~\cite{Beyler:2009}.

Another landmark document regarding the state of forensic science in the United States was published in 2009.  The 328 page report, entitled {\em Strengthening Forensic Science in the United States: A Path Forward}, by the U.S.~National Academy of Sciences, reviewed 13 different forensic science disciplines including biological evidence, friction ridge patterns, tool mark and firearms identification, and analysis of explosives evidence and fire debris.  The review of the analysis of explosives evidence and fire debris is covered in three pages of the report.  The summary assessment of analysis of explosives evidence is positive, ``the scientific foundations exist to support the analysis of explosions, because such analysis is based primarily on well-established chemistry~''~\cite{Forensic:2009}.  The complete summary assessment for fire debris analysis is included here: ``By contrast, much more research is needed on the natural variability of burn patterns and damage characteristics and how they are affected by the presence of various accelerants.  Despite the paucity of research, some arson investigators continue to make determinations about whether or not a particular fire was set.  However, according to testimony presented to the committee, many of the rules of thumb, that were typically assumed to indicate that an accelerant was used (e.g. ``alligatoring'' of wood, specific char patterns) have been shown not to be true.  Experiments should be designed to put arson investigations on a more solid scientific footing .''
This study is in response to the assessment above, specifically regarding the natural variability of fire patterns.

@@

The majority of intentional fires in structures have occurred in residential occupancies, therefore this study focused on fire patterns on painted gypsum wallboard which is a common interior finish for residential occupancies.  A common height for a residential occupancies is 2.4~m.  With a wall height of 2.4~m, these experiments used small fires with flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall would be representative of a pre-flashover fire pattern.

@@

Replicate experiments exposing gypsum wallboard to the fires were conducted in order to examine the repeatability of the fire patterns.  Correlations between the fire characteristics and the fire patterns were be examined.

The last component of this study examines the capability of using numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, as well as bench scale and full scale fire experiments conducted in this study.

@@

This study includes several series of real-scale, replicate fire experiments to examine the repeatability of pre-flashover fire patterns and, ultimately, examine the ability of simple correlations and the Fire Dynamic Simulator to recreate characteristics of the fire patterns.

@@

The most important measurements for the field investigator are the visual measurements that can be made at the scene or determined from scene photographs. Several methods of imaging were used for these experiments: still photography, digital video, and infrared video.  Computer based imaging techniques, such as those contained in the Image Stream program suite, were examined and adapted to analyze the flame height of the source fires~\cite{Nokes:2011}.

Statistical methods were used to guide the wall fire and corner fire experiments.  The analysis ensured that the appropriate number of replicate experiments were conducted to provide an assessment of burn pattern repeatability within a given level of uncertainty.  Since the coefficient of variation (COV) may be different for a given type of experiment, the number of replicates required would be larger for experiments with a larger COV in order obtain a valid assessment of repeatability.

@@

The fires were characterized in terms of heat release rate, mass loss rate, fire plume temperatures and emitted heat flux. Videos were recorded for each experiment.  This ``field'' data from the videos was  used to determine the flame heights. Experiments were conducted with three different fuel types: a gas fueled mass flow controlled burner, a liquid fueled burner, and a solid fuel.

@@

A free standing, steel frame with non-combustible, calcium silicate faced wall was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved closer toward the wall until the burner was against the wall.

@@

A free standing, steel frame with a panel of gypsum wallboard attached to the surface was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was moved closer toward the wall until the burner was against the wall.  If the results indicated that no burn pattern would result from a given combination of fuel and position, then that fuel and position combination were eliminated from the remainder of the test matrix.

@@

The first set of fire pattern experiments on painted gypsum wallboard, without potentially intrusive instrumentation, was conducted to examine the impact of the construction on the wall. Three different types of wall construction were examined to determine if the amount of insulation behind the exposed gypsum wallboard changed the fire pattern by a significant amount.  The results would then be used to plan the wall construction of the remaining full- scale experiments.

@@

A series of experiments were conducted with samples of painted gypsum wallboard to determine the repeatability of a ``critical heat flux'' that results in a clear line of demarcation, a ``burn/no burn'' line.  The flooring radiant panel, ASTM~E 648, and a modified radiant panel apparatus based on ASTM~E 162 were used to examine this value~\cite{ASTM_E648},~\cite{ASTM_E162}.

@@

Cone calorimeter experiments were conducted, based on ASTM~E 1354, to estimate the ignition temperature, the time to ignition and the heat release rate from samples of painted gypsum wallboard under thermal flux conditions ranging from 25~kW/m$^2$ to 75~kW/m$^2$ ~\cite{ASTM_E1354}.

@@

The ability to recreate or simulate a fire pattern or certain characteristics of a fire pattern was examined with a variety of tools.  The first characteristic examined was the ability to determine, based on the calculated heat flux to the wall, if a ``visible or measurable physical change'' would occur to the painted paper surface of the gypsum wallboard.  The next characteristic of the fire pattern that was examined was the height.  The height of the fire pattern was then compared to the predictive results from flame height correlations. Heskestad developed an empirical method for calculating the mean visible flame height for fuels which do not have substantial in-depth combustion~\cite{Beyler:1986,Heskestad:SFPE}. This correlation was compared with the flame height results of the source fires and the instrumented wall experiments.

FDS~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide} were used to examine the ability to numerically reproduce the source fires and the instrumented wall experiments. These simulation tools provide results in the form of visual field data.  In other words they provide results in the same format that the investigator would see or document the fire pattern.  Comparison between the experiments and the model provide valuable validation data for the model.

@@

Point measurements were reduced and compared with values from the literature on fire plumes and their interaction with vertical surfaces.  The heat release rates, heat flux measurements, and flame and surface temperatures were used as input for heat transfer analysis to develop empirical coefficients for the specific experimental arrangements, which would be useful in the model recreations.
The key measurements for repeatability of the fire patterns on the gypsum wallboard are height, maximum width, shape, and area.  The two significant outcomes will be the repeatability of the fire patterns themselves and the ability to numerically recreate key measures of the pattern.

@@

N.B. This section may be considered atypical for an engineering thesis but it is included here for two reasons.  Firstly, to provide a narrative description for readers that may be unfamiliar with how fire patterns are routinely relied on in fire investigations.  The second is to put this work into an applied context that is familiar to fire investigators that may be unaccustomed to reading detailed research as found in this thesis.

>>>>

This section reviews the fire problem to demonstrate the need and value to society that fire investigation can provide. Following the review, a brief presentation on the analysis  of fire patterns, the use of analytical tools, and the current challenges to the field of fire forensics is given because they are the reason for this study.  At the end of this chapter, an example of a fire incident is presented. In Chapter 9, we will return to the example again.

\subsection{The Need for Fire Investigation}
Unwanted fires result in loss of life and property, and can create an adverse economic impact on a community.  Recent findings from an archaeological site in South Africa indicate mankind has productively used fire for approximately 1 million years~\cite{Berna:2012}.  This is 300,000 years earlier than previously thought.  Despite mankind's extensive history with fire, according to the World Health Organization an estimated 265,000 people die each year as a result of burn injuries~\cite{WHO:2014}.  Based on an estimated global population of 7.3 billion people, the death rate due to burns from all sources would be 3.6 per 100,000 people.  The World Fire Statistics Center collects data from European, North American, and a few Asia-Oceania countries.  Within this group of countries the fire deaths per 100,000 people range from 0.02 (Singapore) to 2.03 (Finland)~\cite{Climate:2014}. There are a significant number of factors that determine the variability of fire death rates between nations, socio-economic conditions being a leading factor~\cite{WHO:2014}.

In the United States, the loss of life due to unwanted fire has been decreasing steadily since the 1970’s.  (This trend does not include loss of life as a result of the terrorist attack on September 11, 2001.)  The estimated number of fire deaths in the U.S. in 1975 was 8,100, or 3.75 deaths per 100,000 people~\cite{America_Burning_Revisited}.  In 2013, there were 3,240 civilian fire fatalities, or 1.02 deaths per 100,000 people~\cite{Karter:2014}.  During this same period, the number of fires in the United States has decreased from 2.5 million fires to 1.24 million fires.  There are many reasons that led to this decrease in loss of life due to fire in the United States.  One example is research and data that enabled the development of product safety standards, improved building codes, improved fire safety systems, and fire prevention programs.  None of this happened without the constant effort of agencies at the federal, state, and local levels working with professional fire organizations in order to deliver the five E’s of fire prevention: engineering, education, enforcement, economic Incentive, and emergency Response~\cite{FEMA:2013}.

@@

According to the NFPA, there were approximately 282,600 intentionally set fires each year from 2007 through 2011 in the United States. The average annual loss totals for that period were approximately 420 civilian deaths and 1,360 civilian injuries. More than \$1.3 billion dollars (U.S.) was lost per year due to direct property damage. Structure fires represent less than 20\% of intentionally set fires.  However, fires involving structures account for more than 84\% of the civilian deaths, injuries and property loss from all intentional fires.  Approximately two-thirds of the intentional structure fires occurred in occupied and operating residential occupancies.  As a result, 95\% of the civilian deaths and 86\% of the civilian injuries attributed to intentional fires occurred in residential occupancies~\cite{Campbell:2014}.

A subset of the intentional fires are considered arson.  The U.S. Federal Bureau of Investigation (FBI) Uniform Crime Reporting (UCR) Program defines arson as any willful or malicious burning or attempting to burn, with or without intent to defraud, a dwelling house, public building, motor vehicle or aircraft, personal property of another, etc.~\cite{Crime:2010}.  The national rate of arson fire cases cleared by arrest or exceptional means was less than 20\%.  Of those cases, approximately 46\% of the suspects were prosecuted~\cite{Campbell:2014}.  Hence less than 10\% of the arson cases resulted in successful prosecution.

The statistics cited above are believed to be the best available in the United States, which maintains perhaps one of the better fire incident databases in the world.  However, an investigative report by Icove and Hargrove indicates arson fires are underreported to the U.S. National Fire Incident Reporting System (NFIRS).  The investigation provided examples from several major cities, such as New York City. In 2011, 11 arson fires in buildings were reported to NFIRS, while the Fire Department of the City of New York's Bureau of Fire Investigation recorded 1,347 arson fires in buildings that year.  Similar discrepancies were found in eight other large U.S. cities.  On average, three-fourths of the arsons uncovered by city fire investigators went unreported to NFIRS~\cite{Icove_2014}.  In other words, the number of arson-related incidents is likely to be under represented in the fire incident data.

\subsection{The Use of Fire Patterns in Fire Investigation}
Fires are investigated in order to determine their origin and cause.  In cases of arson, there is an additional effort to determine who was responsible for setting the fire.  This investigation process must follow the scientific method, as documented in NFPA 921~\cite{NFPA:921}.  There are numerous textbooks on the subject of conducting a fire investigation~\cite{Almirall:2004,Fire_Investigation,DeHaan:2012,Icove:2013,Lentini:2006,Noon:1995}. The NFPA guide and the texts provide information how to collect data from the scene, analyze the data, and develop a hypothesis about the fire.

Patterns produced by the fire are in many cases a significant portion of the data collected and are analyzed at the scene to determine the area of origin.  One of the basic methods of documenting the fire scene is to photograph fire patterns.  As noted by Icove, DeHaan, and Haynes,  ``the ability to document and interpret fire patterns accurately is essential to investigators reconstructing fire scenes''~\cite{Icove:2013}.  A fire pattern is defined in NFPA~921 as, ``the visible or measurable physical change, or identifiable shapes, formed by a fire effect or group of fire effects''~\cite{NFPA:921}.

DeHaan~\cite{DeHaan:2012} categorizes the fire effects that form patterns as follows:
\begin{enumerate}
\item Surface deposits: no irreversible effect on surface
\item Surface thermal effects: physical change such as discoloration or melting
\item Charring: evidence of surface burning
\item Penetration: charring below the surface
\item Consumption: loss of surface material, charring throughout
\end{enumerate}

Lentini points out that most fire patterns are generated and later observed on two-dimensional surfaces at the place where those surfaces intersect with the three-dimensional fire. Various types of fire patterns, such as V-shaped, hour-glass, and inverted cone, have come from common observation at actual fire scenes~\cite{Lentini:2006}.  As a result, the observations are typically qualitative in nature.

Previous fire pattern research by the National Institute of Justice (NIJ), the National Institute of Standards and Technology (NIST), and the United States Fire Administration (USFA) has shown that fire patterns provide data useful for the determination of fire origin.  The reports noted the impact of ventilation on the development of the burn patterns~\cite{Shanley:1997,Putorti:1997}. A large number of other factors affect the formation of these patterns: burn time, heat release rate of the fire source, fire exposure, target fuel composition, adjacent fuel(s), and compartmentalization, to name a few. Given the limited number of experiments in the literature and the large number of variables, it has been difficult to identify a cause and effect relationship between fire scenarios and the resulting patterns, based solely on the research.

\subsection{The Use of Fire Models in Fire Investigation}
Fire models are another tool that may be used by a fire investigation team to gain insight into the growth and development of a fire and to test the investigators' hypothesis on origin and cause~\cite{Sutula:2001}.  Many theoretical and empirical correlations and computational fire models have been developed for use in fire protection design, fire hazard analysis, and fire investigation.  A collection of the widely used correlations can be found in the U.S. Nuclear Regulatory Commission’s (NRC) Fire Dynamics Tools (FDTs)~\cite{NUREG_1805}. These spreadsheets incorporate references to the source documents and experimental work upon which the correlations are based. NRC tasked NIST with conducting a verification and validation study of these correlations. Overholt developed the NIST verification and validation study~\cite{Overholt:2014}. Not all of the correlations in the FDTs were included in the study, because a sufficient amount of reliable data was not available to conduct a proper validation.  The correlations for flame height and time to ignition were not included in the validation study.

@@

Of the active field models, FDS version 6 has one of the most rigorous and publicly accessible verification and validation programs~\cite{FDS_Verification_Guide,FDS_Validation_Guide}. More than 45 experimental data sets are used to quantify the prediction capabilities of the model.  Results are presented in terms of model bias and variation to inform users of whether particular quantities (temperature, heat flux, pressure, etc.) are over- or under-predicted compared to experimental data.

In order for fire models to be demonstrated as reliable and suitable for use in court, studies such as the one conducted here must be conducted to provide the scientific basis for the use of fire models for re-construction.  These research results may enable fire investigation teams to have a better understanding of the fire scene and enable improved support of their hypothesis on the area of origin based on fire patterns.  This idea was discussed when the Fire Protection Research Foundation convened a Research Advisory Council on Post-fire Analysis in 2002.  Recommendations for research and development in their white paper included, ``advance the capabilities of computational fluid dynamics fire modeling, particularly as applied to fire ignition scenarios and fire pattern development and interpretation''~\cite{RAC:2002}.

Because fire pattern interpretation and analysis is a critical step in most fire investigations, it is surprising to find the limited amount of research and analysis tools directly related to this practice.  Gorbett recently conducted a review of fire pattern research that has been conducted over the last 80 years. The review points out many gaps, especially with how fire patterns are used to determine the area of origin.  A key limitation is the assumption by many of the researchers and text authors that investigators are able to visibly assess varying degrees of fire damage and then determine the area of origin~\cite{Gorbett_2015}. Studies have shown that investigators have not been able to use only the location of the fire patterns as means of reliably determining the area of origin.  Some amount of analysis and fire dynamics knowledge is required~\cite{Carmen_2008},~\cite{Tinsley_2013}.

\subsection{Fire Investigation Under Review}
The fire investigation community recognized a need to improve the science and practice of fire investigation.  One of the manifestations of the efforts to improve the practice was the development of NFPA~921, {\em Guide for Fire and Explosion Investigations}~\cite{NFPA:921}.  The first edition of NFPA~921 was issued in 1992.  As provided by the document scope, it ``is designed to assist individuals who are charged with the responsibility of investigating and analyzing fire and explosion incidents and rendering opinions as to the origin, cause, responsibility, or prevention of such incidents, and the damage and injuries which arise from such incidents.'' The document’s purpose includes the goal to be ``a model for the advancement and practice of fire and explosion investigation, fire science, and methodology.'' However, NFPA~921 can only incorporate the fire science and research results that have been produced.  Although NFPA~921 is a guide as opposed to a standard, it is considered the standard of care for the fire investigation community in countries that recognize and use NFPA standards and guides.

As a result of the review of several arson cases, most notably the cases of Ernest Ray Willis and Cameron Todd Willingham, the practice of fire investigation as a forensic science was called into question.  Beyler conducted a review of both cases for the Texas Forensic Science Commission~\cite{Beyler:2009}.  Both fires occurred prior to the release of NFPA~921 in 1992.  The Beyler analysis, issued in August of 2009, concluded that the investigations of both cases ``did not comport with either the modern standard of care expressed by NFPA 921, or the standard of care expressed by fire investigation texts and papers in the period 1980{\textendash}1992.'' The fire investigators had a poor understanding of science, and their findings of arson could not be sustained~\cite{Beyler:2009}.

Another landmark document regarding the state of forensic science in the United States was published in 2009.  The 328 page report, entitled {\em Strengthening Forensic Science in the United States: A Path Forward}, by the U.S.~National Academy of Sciences, reviewed 13 different forensic science disciplines including biological evidence, friction ridge patterns, tool mark and firearms identification, and analysis of explosives evidence and fire debris.  The review of the analysis of explosives evidence and fire debris is covered in three pages of the report.  The summary assessment of analysis of explosives evidence is positive, ``the scientific foundations exist to support the analysis of explosions, because such analysis is based primarily on well-established chemistry''~\cite{Forensic:2009}.  The complete summary assessment for fire debris analysis states: ``By contrast, much more research is needed on the natural variability of burn patterns and damage characteristics and how they are affected by the presence of various accelerants.  Despite the paucity of research, some arson investigators continue to make determinations about whether or not a particular fire was set.  However, according to testimony presented to the committee, many of the rules of thumb that were typically assumed to indicate that an accelerant was used (e.g., alligatoring of wood, specific char patterns) have been shown not to be true.  Experiments should be designed to put arson investigations on a more solid scientific footing.''
This study is in response to the assessment above, specifically regarding the natural variability of fire patterns.

@@

The majority of intentional fires in structures have occurred in residential occupancies, so this study focused on fire patterns on painted gypsum wallboard, which is a common interior finish for residential occupancies.  A common height for a residential occupancies is 2.4~m.  With a wall height of 2.4~m, these experiments used small fires with flame heights of approximately 1.2~m.  The resulting patterns on the gypsum board wall were be representative of a pre-flashover fire pattern.

@@

Replicate experiments exposing gypsum wallboard to the fires were conducted in order to examine the repeatability of the fire patterns.  Correlations between the fire characteristics and the fire patterns were examined.

The last component of this study examined the capability of using numerical models to simulate the fires and the resulting fire patterns based on input data collected from engineering reference sources, as well as bench scale and full scale fire experiments conducted in this study.

@@

This study included several series of real-scale, replicate fire experiments to examine the repeatability of pre-flashover fire patterns and, ultimately, examine the ability of simple correlations and the Fire Dynamic Simulator to recreate characteristics of the fire patterns.

@@

The most important measurements for the field investigator are the visual measurements that can be made at the scene or determined from scene photographs. Several methods of imaging were used for these experiments: still photography, digital video, and infrared video.  Computer-based imaging techniques, such as those contained in the Image Stream program suite, were examined and adapted to analyze the flame height of the source fires~\cite{Nokes:2011}.

Statistical methods were used to guide the wall fire and corner fire experiments.  The analysis ensured that the appropriate number of replicate experiments were conducted to provide an assessment of burn pattern repeatability within a given level of uncertainty.  Because the coefficient of variation (COV) may be different for a given type of experiment, the number of replicates required would be larger for experiments with a larger COV in order obtain a valid assessment of repeatability.

@@

The fires were characterized in terms of heat release rate, mass loss rate, fire plume temperatures, and emitted heat flux. Videos were recorded for each experiment.  This ``field'' data from the videos was  used to determine the flame heights. Experiments were conducted with three different fuel types: a gas fueled mass flow controlled burner, a liquid fueled burner, and a solid fuel.

@@

A free standing, steel frame with a non-combustible, calcium silicate-faced wall was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was then moved toward the wall until the burner was against the wall.

@@

A free standing, steel frame with a panel of gypsum wallboard attached to the surface was instrumented to measure the thermal exposure of the wall from the fire in terms of temperature and heat flux.  The fires were initially placed at predetermined, fixed positions away from the wall. The burner was then moved toward the wall until the burner was against the wall.  If the results indicated that no burn pattern would result from a given combination of fuel and position, then that fuel and position combination were eliminated from the remainder of the test matrix.

@@

The first set of fire pattern experiments on painted gypsum wallboard, without potentially intrusive instrumentation, was conducted to examine the impact of the construction on the wall. Three different types of wall construction were examined to determine if the amount of insulation behind the exposed gypsum wallboard changed the fire pattern by a significant amount.  The results were used to plan the wall construction of the remaining full- scale experiments.

@@

A series of experiments were conducted with samples of painted gypsum wallboard to determine the repeatability of a critical heat flux that results in a clear line of demarcation, a burn/no burn line.  The flooring radiant panel, ASTM~E 648, and a modified radiant panel apparatus based on ASTM~E 162 were used to examine this value~\cite{ASTM_E648},~\cite{ASTM_E162}.

@@

Cone calorimeter experiments were conducted, based on ASTM~E 1354, to estimate the ignition temperature, the time to ignition, and the heat release rate from samples of painted gypsum wallboard under thermal flux conditions ranging from 25~kW/m$^2$ to 75~kW/m$^2$ ~\cite{ASTM_E1354}.

@@

The ability to recreate or simulate a fire pattern or certain characteristics of a fire pattern was examined with a variety of tools.  The first characteristic examined was the ability to determine, based on the calculated heat flux to the wall, if a visible or measurable physical change would occur to the painted paper surface of the gypsum wallboard.  The next characteristic of the fire pattern examined was the height.  The height of the fire pattern was then compared to the predictive results from flame height correlations. Heskestad developed an empirical method for calculating the mean visible flame height for fuels that do not have substantial in-depth combustion~\cite{Beyler:1986,Heskestad:SFPE}. This correlation was compared with the flame height results of the source fires and the instrumented wall experiments.

FDS~\cite{FDS_Users_Guide} and Smokeview~\cite{Smokeview_Users_Guide} were used to examine the ability to numerically reproduce the source fires and the instrumented wall experiments. These simulation tools provide results in the form of visual field data.  In other words, they provide results in the same format an investigator would see or use to document the fire pattern.  Comparison between the experiments and the model provide valuable validation data for the model.

@@

Point measurements were reduced and compared with values from the literature on fire plumes and their interaction with vertical surfaces.  The heat release rates, heat flux measurements, and flame and surface temperatures were used as input for heat transfer analysis to develop empirical coefficients for the specific experimental arrangements, which would be useful in the model recreations.
The key measurements for repeatability of the fire patterns on the gypsum wallboard are height, maximum width, shape, and area.  The two significant outcomes would be the repeatability of the fire patterns themselves, and the ability to numerically recreate key measures of the pattern.

@@

N.B. This section may be considered atypical for an engineering thesis, but it is included here for two reasons.  First, to provide a narrative description for readers that may be unfamiliar with how fire patterns are routinely relied on in fire investigations.  Second, is to put this work into an applied context that is familiar to fire investigators who may be unaccustomed to reading detailed research.

For the moment let's step into the shoes of a fire investigator and examine a potential investigation scenario.  It is 8:30 on a cool fall morning and you are called to a fire scene.  As you prepare to leave for the fire scene, you begin to collect information on the fire as you listen to your department radio.  You have the address.  You start to develop an image in your mind about the location of the home involved in the fire. The fire occurred in area of older town homes.  The socio-economics of the neighborhood are such that the people who live there have many life struggles. En route you learn the fire department rescued a mother and child from the home and that they are being transported to the hospital.  The husband and father also lives at that address but is unaccounted for at this time.  Upon arrival you have a conversation with the police and fire officers on scene.  You are provided with information regarding what the first firefighters saw on scene: ``All of the doors and windows were closed, and smoke was leaking from around the front door and living room window.  The front door was forced and a hose stream was applied to the fire in the living room.  Fire knockdown was rapid.  The mother and child were rescued from a bedroom upstairs.'' You thank the fire officer and get contact information for a follow-up if necessary.  As you wait for the fire department to complete their operations inside, the police on scene let you know the address is familiar to them for domestic disturbance calls for the husband and wife fighting, but no charges have been filed.  They are looking for the husband and have sent a patrol car to his place of work.

As you begin to examine the scene, you walk around the exterior and note that on the front of the end unit townhouse, there is fire damage on the exterior above the front door opening that extends across the ceiling of the porch.  There is some thermal damage and smoke deposits on the front siding above the porch.  The outer pane of the living room window is, for the most part, still intact. You notice that three of the four windows on the second floor have been removed.  On side D of the town home, there are no openings and no smoke or fire damage is present.  On the rear or side C of the house, the door and windows are intact but open.  There are no smoke stains and no thermal damage around any of the openings on the rear. No sign of smoke alarms in the home.
